### 2022-06-04 omicron-generation.tex
update

\begin{flushleft}{
	\Large
	\textbf\newline{
		Inferring the differences in incubation-period and generation-interval distributions of the Delta and Omicron variants of SARS-CoV-2
	}
}
\newline
\\
Sang Woo Park\textsuperscript{1,*},
Kaiyuan Sun\textsuperscript{2},
Sam Abbott\textsuperscript{3},
Ron Sender\textsuperscript{4},
Yinon Bar-on\textsuperscript{4},
Joshua Weitz\textsuperscript{5,6,7},
Sebastian Funk\textsuperscript{3,8}, 
Bryan T. Grenfell\textsuperscript{1,9},
Jantien A Backer\textsuperscript{10},
Jacco Wallinga\textsuperscript{10,11},
Cecile Viboud\textsuperscript{2},
Jonathan Dushoff\textsuperscript{12,13,14}
\\
\bigskip
\textbf{1} Department of Ecology and Evolutionary Biology, Princeton University, Princeton, NJ, USA
\\
\textbf{2} Division of International Epidemiology and Population Studies, Fogarty International Center, 20 National Institutes of Health, Bethesda, MD, USA
\\
\textbf{3} Centre for the Mathematical Modelling of Infectious Diseases, London School of Hygiene \& Tropical Medicine, London, UK
\\
\textbf{4} Weizmann Institute of Science, Rehovot, Israel
\\
\textbf{5} School of Biological Sciences, Georgia Institute of Technology, Atlanta, GA, USA
\\
\textbf{6} School of Physics, Georgia Institute of Technology, Atlanta, GA, USA
\\
\textbf{7} Institut de Biologie, \'{E}cole Normale Sup\'{e}rieure, Paris, France
\\
\textbf{8} Department of Infectious Disease Epidemiology, London School of Hygiene \& Tropical Medicine, Keppel Street, London, UK
\\
\textbf{9} Princeton School of Public and International Affairs, Princeton University, Princeton, NJ, USA
\\
\textbf{10} Center for Infectious Disease Control, National Institute for Public Health and the Environment (RIVM), Bilthoven, The Netherlands
\\
\textbf{11} Department of Biomedical Data Sciences, Leiden University Medical Center, Leiden, The Netherlands
\\
\textbf{12} Department of Biology, McMaster University, Hamilton, ON, Canada
\\
\textbf{13} Department of Mathematics and Statistics, McMaster University, Hamilton, ON, Canada
\\
\textbf{14} M.\,G.\,DeGroote Institute for Infectious Disease Research, McMaster University, Hamilton, ON, Canada
\\
\bigskip

@@

Here, we re-analyze serial-interval data collected by \citep{backer2021omicron}, representing within- and between-household transmissions of the Delta and Omicron variants from the Netherlands between 13 and 26 December 2021.
The study found shorter mean within-household serial intervals (3.5 vs 4.1 days) and mean incubation periods (3.2 vs 4.4 days) for transmission pairs with S-gene target failure (mostly Omicron during the study period) than without (mostly Delta), but did not consider dynamical biases caused by growth-rate differences in their inference:
during this period, the Omicron cases were increasing, whereas the Delta cases were decreasing.
We take the epidemiological context in the Netherlands during the study period into account to provide corrected estimates for the incubation periods and generation-interval distributions of the Delta and Omicron variants.
We show that using up-to-date generation-interval distributions is critical to accurately estimating the reproduction advantage (i.e., the ratio between the reproduction numbers of the invading and resident variants) of emerging SARS-CoV-2 variants.

@@

Publicly available data are aggregated by the length of the serial interval in days and do not include additional individual-level information, such as exposure dates, symptom onset dates, or age.
The original article presented serial-interval estimates stratified by the vaccination status in supplementary materials, but stratified data are not publicly available; we rely on publicly available data to keep the analysis simple and to focus on the qualitative effects of dynamical biases.
The aggregated data consists of 2529 transmission pairs and are further stratified by the presence of S gene target failure (SGTF), week of infectors' symptom onset date (week 50, 13--19 December 2021, and week 51, 20--26 December 2021), and the type of transmission (within- or between-household transmission).
In the main text, we combine data from weeks 50 and 51 of 2021 (13--26 December) and present a stratified analysis in Supplementary Material.
For simplicity, we refer to transmission pairs with and without SGTFs as Omicron and Delta transmission pairs, respectively.
Incubation period data are not publicly available with the original article; instead, we rely on previous estimates \cite{backer2021omicron} to derive growth-rate-adjusted incubation-period distributions.

@@

In order to accurately estimate incubation-period and generation-interval distributions of the Delta and Omicron variants, we have to take their epidemiological dynamics---in particular, the differences in their growth rates---into account.
To estimate the growth rates differences of the Delta and Omicron variants, we first estimate the number of COVID-19 cases caused by each variant by multiplying reported weekly numbers of cases by the proportion of Delta and Omicron variants detected---we use weekly time series to smooth over patterns of testing and reporting within each week. 
We note that the proportion of Delta and Omicron variants detected is reported by the date of sampling, whereas the case data are reported by the date of reports, meaning that there is some delay between the two data sets (typically around 2 days).
For simplicity, we do not account for this delay, which in turn affects our growth-rate estimates; 
instead, we later perform sensitivity analysis to assess how growth rates affect the inferences of the incubation-period and generation-interval distributions.
We also do not account for uncertainties around the estimates of the proportion of each variant---almost 2000 samples were tested on each week between the week of November 28, 2021, and the week of January 23, 2022, making this uncertainty due to sample size negligible.

We then fit a generalized additive model \citep{wood2001mgcv} to the logged weekly case estimates to obtain smooth trajectories for case time series.
More specifically, we model the logged weekly numbers of cases caused by each variant as a function of time using a shrinkage version of a cubic spline with restricted maximum likelihood (specified as \texttt{s(time, bs="cs")} using the \texttt{MGCV} \texttt{R} package). 
We also assume normally distributed errors around the logged weekly cases.
In principle, using negative-binomial likelihoods with a log link function is more appropriate for modeling count data, especially when data are overdispersed \citep{ver2007quasi};
here, we use Gaussian likelihoods on logged cases because the inferred numbers of cases caused by each variant are not counts.
Finally, we take the derivative of the predicted logged numbers of cases caused by each variant to obtain time-varying growth rate estimates.

@@

In particular, when incidence of infection is growing exponentially, we are more likely to observe backward incubation periods that are shorter than the corresponding forward incubation periods because there will be relatively more individuals who were infected recently.
Assuming that incidence of infection is changing exponentially at a constant rate $r$ across the study period, the backward incubation-period distribution $b_I(\tau)$ corresponds to:
\begin{equation}
b_I(\tau) =\frac{ \exp(-r\tau) f_I(\tau)}{\int_0^\infty \exp(-r x) f_I(x)  \dx}.
\label{eq:backward}
\end{equation}
Therefore, the backward incubation-period distribution $b_I(\tau)$ gives a biased estimate of the corresponding forward distribution $f_I(\tau)$.
The method of \cite{backer2020incubation} starts from observed symptom onsets, and estimates the backward incubation-period distribution.

@@

\begin{figure}[!tp]
\includegraphics[width=\textwidth]{figure/figure_diagram.ggp.pdf}
\caption{
\textbf{Schematic diagrams of serial and generation intervals from symptom- and infection-based infector cohorts.}
(A) Serial intervals are typically measured from the cohort of infectors who develop symptoms at the same time. 
In this case, infectors will have shorter incubation periods than their infectees on average; the corresponding generation intervals will be also short because infectors with short incubation periods will transmit earlier.
(B) Generation intervals for the cohort of infectors who are infected at the same time will have unbiased incubation periods.
\label{fig:diagram}
}
\end{figure}

@@

When an epidemic is growing exponentially, there are two opposing effects affecting the relationship between the mean serial and generation interval.
First, infectors in a given cohort are more likely to have shorter incubation periods than their infectees on average, $\mathbb E[\tau_{i, 1}] < \mathbb E[\tau_{i, 2}]$, causing the mean forward serial interval to be longer than the mean symptom-based generation interval ($\mathbb E[\tau_s] > \mathbb E[\tau_{g, \mathrm{symp}}]$).
Second, the mean symptom-based generation interval will be shorter than the mean infection-based generation interval: $E[\tau_{g, \mathrm{inf}}] > \mathbb E[\tau_{g, \mathrm{symp}}]$.
Therefore, the difference between the mean serial interval and the mean infection-based generation interval is difficult to predict in general; in most cases, however, we expect the former effect to dominate, causing the mean serial interval to be longer than the mean infection-based generation interval: $\mathbb E[\tau_s] > E[\tau_{g, \mathrm{inf}}]$ \citep{park2021forward}.
Earlier work on serial-interval distributions neglected dynamical biases in the incubation periods of the infectors \citep{svensson2007note,britton2019estimation}, which allowed the authors to conclude that the mean generation and serial intervals are identical.
For simplicity, we will use the term ``forward generation-interval'' to refer to the infection-based generation-interval distribution (measured from a cohort of infectors who were infected at the same infection time, \fref{diagram}B), and drop the subscript $\mathrm{inf}$.

Assuming that the incidence of infection is changing exponentially at a constant rate $r$, the forward serial-interval distribution for a cohort of infectors who developed symptoms at the same time $t$ is expected to remain unchanged across $t$ \citep{park2021forward}. 
Then, we can focus on the forward serial-interval distribution at $t=0$, which in turn allows us to reparameterize the incubation-period and generation-interval distributions in terms of the infection time of the infector $\alpha_1 < 0$ and of the infectee $\alpha_2 > \alpha_1$. 
Under this parameterization, for a given length of a serial interval $\tau$, we can rewrite the incubation period of the infector as $-\alpha_1$; the generation interval as $\alpha_2 - \alpha_1$; and the incubation period of the infectee as $\tau - \alpha_2$.
Then, the forward serial-interval distribution $f_S(\tau)$ for a cohort of infectors who developed symptoms at time $t = 0$ can be expressed in terms of three distributions (\eref{serial}): the backward incubation-period distribution of the infector $b_I(-\alpha_1)$, the forward generation-interval distribution conditional on a known value of the incubation period $x$, $f_{G|I}(\alpha_2-\alpha_1|-\alpha_1)$, and the forward incubation-period distribution of the infectee $f_I(\tau - \alpha_2)$.
Integrating across infection time of the infector $\alpha_1 < 0$ and of the infectee $\alpha_2 > \alpha_1$ and rewriting the backward incubation-period distribution $b_I(-\alpha_1)$ in terms of the forward distribution, we obtain \citep{park2021forward}:
\begin{equation}
f_S(\tau) = \frac{1}{\phi} \int_{-\infty}^0\int_{\alpha_1}^\tau \exp(r \alpha_1) f_{G|I}(\alpha_2 - \alpha_1|- \alpha_1) f_I(- \alpha_1) f_I(\tau - \alpha_2) \dd \alpha_2 \dd \alpha_1,
\label{eq:forwardserial}
\end{equation}
where $\phi$ is a normalization constant chosen so that $\int f_S(x) \dx = 1$.
As discussed earlier, this method assumes that the incidence is changing exponentially at a constant rate $r$ across the study period.
As we show later in the results section, the exponential growth rate changes over the study period, including weeks 50 and 51 (13--26 December 2021);
for illustrative purpose, we choose values of $r$ that represent the dynamics of Delta and Omicron infections during this period and repeat the analysis across plausible ranges of $r$ (discussed later in detail).

@@

We model the forward incubation-period $f_I(\tau)$ and generation-interval $f_G(\tau)$ distributions using a bivariate lognormal distribution.
The joint distribution is parameterized by log means, $\mu_I$ and $\mu_G$, log variances, $\sigma_I^2$ and $\sigma_G^2$, and the correlation coefficient on a log scale $\rho$.
Thus, the forward generation-interval distribution conditional on the incubation period $f_{G|I}(\tau|\tau_{i,1})$ has a log mean of $\mu_G + \sigma_G \rho (\log(\tau_{i, 1}) - \mu_I)/\sigma_I$ and a log variance of $\sigma_G^2 (1 - \rho^2)$.
For a given value of $r$, we first estimate the forward incubation-period distribution from the backward distribution, previously estimated by \cite{backer2021omicron}, using \eref{forward}.
We then approximate the forward incubation-period distribution with a lognormal distribution by matching the mean and standard deviation---this is also known as the method of moments.
Using this incubation-period distribution, we fit \eref{forwardserial} to the observed serial-interval data by minimizing the negative log-likelihood.
We then calculate the mean forward generation interval using \eref{meangen}.
The 95\% confidence intervals are calculated by taking the estimated variance-covariance matrix for the log-mean and -standard deviation parameters of the log normal distributions and calculating the corresponding variance-covariance for the overall mean using Taylor expansion---this method is also known as the Delta method \citep{oehlert1992note}.
We assume $\rho = 0.75$ throughout based on \cite{sender2021unmitigated}---since we do not have individual-level data on infection and symptom onset times, we expect this parameter to be unidentifiable in practice.
In Supplementary Material, we explore how assumptions about $\rho$ affect inferences of the generation-interval distribution.

@@

\fref{epidemic} summarizes the epidemiological context in the Netherlands during the study period.
The first known Omicron case in the Netherlands was sampled on 19 November 2021 \citep{backer2021omicron}, during a period when COVID-19 incidence was decreasing (\fref{epidemic}A).
As the Omicron variant continued to spread and increase in proportion (\fref{epidemic}B), the number of COVID-19 cases started to increase (\fref{epidemic}A).
Multiplying the proportion of each variant with the number of reported COVID-19 cases further allows us to estimate the epidemiological dynamics of each (\fref{epidemic}C).
The number of COVID-19 cases caused by the Delta variant continued to decrease throughout the study period with time-varying growth rates decreasing from $r \approx -0.01/\mathrm{day}$ to $r \approx -0.09/\mathrm{day}$ by the week of January 16, 2022, and increasing back up to $r \approx -0.04/\mathrm{day}$ by the end of January, 2022 (\fref{epidemic}D).
The number of COVID-19 cases caused by the Omicron variant increased rapidly but decelerated over time with time-varying growth rates decreasing from $r=0.18/\mathrm{day}$ on the week of December 19, 2021, to $r=0.04/\mathrm{day}$ by the end of January, 2022.
These changes in growth rates coincide with the introduction of lockdown on 19 December 2021 \citep{netherlandslockdown} and its relaxation beginning 15 January 2022 \citep{netherlandsrelax,netherlandsopen}.
We note that the growth-rate difference between the Delta and Omicron variants decreased over time.
Hereafter, we use $r=-0.05/\mathrm{day}$ for the Delta variant and $r=0.15/\mathrm{day}$ for the Omicron variant as representative growth rates---these growth rates correspond to the mean growth rates between 1 December 2021 and 2 January 2022, during which the incubation-period data were collected.
We then evaluate the growth-rate effects across $r=-0.1/\mathrm{day}$--$0.0/\mathrm{day}$ for the Delta variant and $r=0.1/\mathrm{day}$--$0.2/\mathrm{day}$ for the Omicron variant as a sensitivity analysis.

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{figure/figure_epidemic.ggp.pdf}
\caption{
\textbf{Epidemic dynamics on the Delta and Omicron variants in the Netherlands between November 2021 and January 2022.}
(A) Daily numbers of reported COVID-19 cases in the Netherlands (points).
The solid line represents the 7-day moving average.
Data are publicly available on \url{https://data.rivm.nl/covid-19/}.
(B) Proportion of SARS-CoV-2 variants detected from the Netherlands. Data are publicly available on \url{https://www.rivm.nl/coronavirus-covid-19/virus/varianten}.
(C) Weekly numbers of COVID-19 cases caused by the Delta (black points) and Omicron (orange triangles) variants are estimated by multiplying the weekly numbers of cases (A) with the proportion of each variant (B).
Solid lines and shaded areas represent fitted lines and corresponding 95\% confidence intervals using generalized additive model.
(D) Estimated growth rates of the Delta (black) and Omicron variants (orange) and their growth-rate differences (purple).
Lines and shaded areas represent medians and corresponding 95\% confidence intervals.
Growth rates are estimated by taking the derivative of the generalized additive model estimates of logged number of cases.
\label{fig:epidemic}
}
\end{figure}

@@

Our corrected estimates of the forward incubation-period distributions further allow us to infer the forward generation-interval distributions.
For illustrative purposes, we first focus on aggregated serial intervals from infectors who developed symptoms during week 50--51 (13--26 December, 2021).
For within-household transmission pairs (\fref{serial}A), the Omicron variant has shorter mean serial interval (3.1 days; 95\% CI: 2.9--3.3 days) than that of the Delta variant (3.7 days; 95\% CI: 3.5--3.8 days).
When we account for growth-rate differences (assuming $r=-0.05/\mathrm{day}$ and $r=0.15/\mathrm{day}$ for the Delta and Omicron variants, respectively), the estimated mean forward generation interval exhibits a slightly larger difference (\fref{serial}B):
3.0 days (95\% CI: 2.7--3.2 days) for the Omicron variant and 3.8 days (95\% CI: 3.7--4.0 days) for the Delta variant.
Across plausible ranges of assumptions about the growth rates of the Delta and Omicron variants, we estimate robust differences in their mean generation intervals (\fref{serial}C--D).
Assuming lower values of the correlation $\rho$ between the incubation period and generation intervals leads to larger differences in the mean generation intervals of the Delta and Omicron variants (Supplementary Figure S1).
In particular, the generation-interval estimates of the Omicron variant are more sensitive to the assumed values of $\rho$ due to faster changes in incidence of infection---for example, changing $\rho$ from 0.85 to 0.5 changes the mean generation-interval estimates for the Omicron variant from 3.1 days (95\% CI: 2.8--3.3 days) to 2.7 days (95\% CI: 2.5--2.9 days).

>>>>

\begin{flushleft}{
	\Large
	\textbf\newline{
		Inferring the differences in incubation-period and generation-interval distributions of the Delta and Omicron variants of SARS-CoV-2
	}
}
\newline
\\
Sang Woo Park\textsuperscript{1,*},
Kaiyuan Sun\textsuperscript{2},
Sam Abbott\textsuperscript{3},
Ron Sender\textsuperscript{4},
Yinon Bar-on\textsuperscript{4},
Joshua Weitz\textsuperscript{5,6,7},
Sebastian Funk\textsuperscript{3,8}, 
Bryan T. Grenfell\textsuperscript{1,9},
Jantien A Backer\textsuperscript{10},
Jacco Wallinga\textsuperscript{10,11},
Cecile Viboud\textsuperscript{2},
Jonathan Dushoff\textsuperscript{12,13,14}
\\
\bigskip
\textbf{1} Department of Ecology and Evolutionary Biology, Princeton University, Princeton, NJ, USA
\\
\textbf{2} Division of International Epidemiology and Population Studies, Fogarty International Center, 20 National Institutes of Health, Bethesda, MD, USA
\\
\textbf{3} Centre for the Mathematical Modelling of Infectious Diseases, London School of Hygiene \& Tropical Medicine, London, UK
\\
\textbf{4} Department of Plant and Environmental Sciences, Weizmann Institute of Science, Rehovot, Israel
\\
\textbf{5} School of Biological Sciences, Georgia Institute of Technology, Atlanta, GA, USA
\\
\textbf{6} School of Physics, Georgia Institute of Technology, Atlanta, GA, USA
\\
\textbf{7} Institut de Biologie, \'{E}cole Normale Sup\'{e}rieure, Paris, France
\\
\textbf{8} Department of Infectious Disease Epidemiology, London School of Hygiene \& Tropical Medicine, Keppel Street, London, UK
\\
\textbf{9} Princeton School of Public and International Affairs, Princeton University, Princeton, NJ, USA
\\
\textbf{10} Center for Infectious Disease Control, National Institute for Public Health and the Environment (RIVM), Bilthoven, The Netherlands
\\
\textbf{11} Department of Biomedical Data Sciences, Leiden University Medical Center, Leiden, The Netherlands
\\
\textbf{12} Department of Biology, McMaster University, Hamilton, ON, Canada
\\
\textbf{13} Department of Mathematics and Statistics, McMaster University, Hamilton, ON, Canada
\\
\textbf{14} M.\,G.\,DeGroote Institute for Infectious Disease Research, McMaster University, Hamilton, ON, Canada
\\
\bigskip

@@

Here, we re-analyze serial-interval data collected by \citep{backer2021omicron}, representing within- and between-household transmissions of the B.1.617.2 (Delta) and B.1.1.529 (Omicron) variants from the Netherlands between 13 and 26 December 2021.
The study found shorter mean within-household serial intervals (3.5 vs 4.1 days) and mean incubation periods (3.2 vs 4.4 days) for transmission pairs with S-gene target failure (mostly Omicron during the study period) than without (mostly Delta), but did not consider dynamical biases caused by growth-rate differences in their inference:
during this period, the incidence of Omicron cases were increasing, whereas the incidence of Delta cases were decreasing.
We take the epidemiological context in the Netherlands during the study period into account to provide corrected estimates for the incubation periods and generation-interval distributions of the Delta and Omicron variants.
We show that using up-to-date generation-interval distributions is critical to accurately estimating the reproduction advantage (i.e., the ratio between the reproduction numbers of the invading and resident variants) of emerging SARS-CoV-2 variants.

@@

Publicly available data are aggregated by the length of the serial interval in days and do not include additional individual-level information, such as exposure dates, symptom onset dates, or age.
The original article presented serial-interval estimates stratified by the vaccination status in supplementary materials, but stratified data are not publicly available; we rely on publicly available data to keep the analysis simple and to focus on the qualitative effects of dynamical biases.
The aggregated data consists of 2529 transmission pairs and are further stratified by the presence of S gene target failure (SGTF), week of infectors' symptom onset date (week 50, 13--19 December 2021, and week 51, 20--26 December 2021), and the type of transmission (within- or between-household transmission).
In the main text, we combine data from weeks 50 and 51 of 2021 (13--26 December) and present a stratified analysis in Supplementary Material.
For simplicity, we refer to transmission pairs with and without SGTFs as Omicron and Delta transmission pairs, respectively.
Incubation period data were originally collected from 513 individuals (consisting of 258 Omicron 165 and 255 Delta cases), with symptom onsets between 1 December 2021 and 2 January 2022;
however, the data are not publicly available with the original article. 
Instead, we rely on previous estimates \cite{backer2021omicron} to derive growth-rate-adjusted incubation-period distributions.

@@

In order to accurately estimate incubation-period and generation-interval distributions of the Delta and Omicron variants, we have to take their epidemiological dynamics---in particular, the differences in their growth rates---into account.
To estimate the growth rates differences of the Delta and Omicron variants, we first estimate the number of COVID-19 cases infected with each variant by multiplying reported weekly numbers of cases by the proportion of Delta and Omicron variants detected---we use weekly time series to smooth over patterns of testing and reporting within each week. 
We note that the proportion of Delta and Omicron variants detected is reported by the date of sampling, whereas the case data are reported by the date of reports, meaning that there is some delay between the two data sets (typically around 2 days).
For simplicity, we do not account for this delay, which in turn affects our growth-rate estimates; 
instead, we later perform sensitivity analysis to assess how growth rates affect the inferences of the incubation-period and generation-interval distributions.
We also do not account for uncertainties around the estimates of the proportion of each variant---almost 2000 samples were tested on each week between the week of November 28, 2021, and the week of January 23, 2022, making this uncertainty due to sample size negligible;
we note that the estimates of the proportion of each variant can be still sensitive to non-randomness of the sampling.

We then fit a generalized additive model \citep{wood2001mgcv} to the logged weekly case estimates to obtain smooth trajectories for case time series.
More specifically, we model the logged weekly numbers of cases infected with each variant as a function of time using a shrinkage version of a cubic spline with restricted maximum likelihood (specified as \texttt{s(time, bs="cs")} using the \texttt{MGCV} \texttt{R} package).
We also assume normally distributed errors around the logged weekly cases.
In principle, using negative-binomial likelihoods with a log link function is more appropriate for modeling count data, especially when data are overdispersed \citep{ver2007quasi};
here, we use Gaussian likelihoods on logged cases because the inferred numbers of cases infected with each variant are not counts.
Finally, we take the derivative of the predicted logged numbers of cases infected with each variant to obtain time-varying growth rate estimates.

@@

In particular, when incidence of infection is growing exponentially, we are more likely to observe backward incubation periods that are shorter than the corresponding forward incubation periods because there will be relatively more individuals who were infected recently.
Assuming that incidence of infection is changing exponentially at a constant rate $r$ across the study period, the backward incubation-period distribution $b_I(\tau)$ corresponds to:
\begin{equation}
b_I(\tau) =\frac{ \exp(-r\tau) f_I(\tau)}{\int_0^\infty \exp(-r x) f_I(x)  \dx},
\label{eq:backward}
\end{equation}
where the denominator is a normalization constant so that $b_I(\tau)$ integrates to 1.
Therefore, the backward incubation-period distribution $b_I(\tau)$ gives a biased estimate of the corresponding forward distribution $f_I(\tau)$.
The method of \cite{backer2020incubation} starts from observed symptom onsets, and estimates the backward incubation-period distribution.

@@

\begin{figure}[!tp]
\includegraphics[width=\textwidth]{figure/figure_diagram.ggp.pdf}
\caption{
\textbf{Schematic diagrams of serial and generation intervals from symptom- and infection-based infector cohorts.}
(A) Forward serial intervals are measured from the cohort of infectors who develop symptoms at the same time. 
In this case, infectors will have shorter incubation periods than their infectees on average; the corresponding generation intervals will be also short because infectors with short incubation periods will transmit earlier.
(B) Generation intervals for the cohort of infectors who are infected at the same time will have unbiased incubation periods.
\label{fig:diagram}
}
\end{figure}

@@

In summary, when an epidemic is growing exponentially, there are two opposing effects affecting the relationship between the mean forward serial and generation interval.
First, infectors who developed symptoms at the same time are more likely to have shorter (backward) incubation periods than the corresponding forward incubation periods of their infectees on average, $\mathbb E[\tau_{i, 1}] < \mathbb E[\tau_{i, 2}]$, causing the mean forward serial interval to be longer than the mean symptom-based generation interval ($\mathbb E[\tau_s] > \mathbb E[\tau_{g, \mathrm{symp}}]$).
Second, the mean symptom-based generation interval will be shorter than the mean infection-based generation interval: $E[\tau_{g, \mathrm{inf}}] > \mathbb E[\tau_{g, \mathrm{symp}}]$ due to correlations between incubation periods and generation intervals.
Therefore, the difference between the mean forward serial interval and the mean infection-based generation interval is difficult to predict in general; in most cases, however, we expect the former effect to dominate, causing the mean forward serial interval to be longer than the mean infection-based generation interval: $\mathbb E[\tau_s] > E[\tau_{g, \mathrm{inf}}]$ \citep{park2021forward}.
Earlier work on serial-interval distributions neglected dynamical biases in the incubation periods of the infectors \citep{svensson2007note,britton2019estimation}, which allowed the authors to conclude that the mean generation and serial intervals are identical.
For simplicity, we will use the term ``forward generation-interval'' to refer to the infection-based generation-interval distribution (measured from a cohort of infectors who were infected at the same infection time, \fref{diagram}B), and drop the subscript $\mathrm{inf}$.

Assuming that the incidence of infection is changing exponentially at a constant rate $r$, the forward serial-interval distribution for a cohort of infectors who developed symptoms at the same time $t$ is expected to remain unchanged across $t$ \citep{park2021forward}. 
Then, we can focus on the forward serial-interval distribution at $t=0$, which in turn allows us to reparameterize the incubation-period and generation-interval distributions in terms of the infection time of the infector $\alpha_1 < 0$ and of the infectee $\alpha_2 > \alpha_1$. 
Under this parameterization, for a given length of a serial interval $\tau$, we can rewrite the incubation period of the infector as $-\alpha_1$; the generation interval as $\alpha_2 - \alpha_1$; and the incubation period of the infectee as $\tau - \alpha_2$.
Then, the forward serial-interval distribution $f_S(\tau)$ for a cohort of infectors who developed symptoms at time $t = 0$ can be expressed in terms of three distributions (\eref{serial}): the backward incubation-period distribution of the infector $b_I(-\alpha_1)$ (taken from \eref{backward}), the forward generation-interval distribution conditional on a known value of the incubation period $x$, $f_{G|I}(\alpha_2-\alpha_1|-\alpha_1)$, and the forward incubation-period distribution of the infectee $f_I(\tau - \alpha_2)$.
Integrating across infection time of the infector $\alpha_1 < 0$ and of the infectee $\alpha_2 > \alpha_1$ and rewriting the backward incubation-period distribution $b_I(-\alpha_1)$ in terms of the forward distribution, we obtain \citep{park2021forward}:
\begin{equation}
f_S(\tau) = \frac{1}{\phi} \int_{-\infty}^0\int_{\alpha_1}^\tau \exp(r \alpha_1) f_{G|I}(\alpha_2 - \alpha_1|- \alpha_1) f_I(- \alpha_1) f_I(\tau - \alpha_2) \dd \alpha_2 \dd \alpha_1,
\label{eq:forwardserial}
\end{equation}
where $\phi$ is a normalization constant chosen so that $\int f_S(x) \dx = 1$.
As discussed earlier, this method assumes that the incidence is changing exponentially at a constant rate $r$ across the study period.
As we show later in the results section, the exponential growth rate changes over the study period, including weeks 50 and 51 (13--26 December 2021);
for illustrative purpose, we choose values of $r$ that represent the dynamics of Delta and Omicron infections during this period and repeat the analysis across plausible ranges of $r$ (discussed later in detail).

@@

We model the forward incubation-period $f_I(\tau)$ and generation-interval $f_G(\tau)$ distributions using a bivariate lognormal distribution.
The joint distribution is parameterized by log means, $\mu_I$ and $\mu_G$, log variances, $\sigma_I^2$ and $\sigma_G^2$, and the correlation coefficient on a log scale $\rho$.
Thus, the forward generation-interval distribution conditional on the incubation period $f_{G|I}(\tau|\tau_{i,1})$ has a log mean of $\mu_G + \sigma_G \rho (\log(\tau_{i, 1}) - \mu_I)/\sigma_I$ and a log variance of $\sigma_G^2 (1 - \rho^2)$.
For a given value of $r$, we first estimate the forward incubation-period distribution from the backward distribution, previously estimated by \cite{backer2021omicron}, using \eref{forward}.
We then approximate the forward incubation-period distribution with a lognormal distribution by matching the mean and standard deviation (also known as the method of moments); we note that we are unable to directly fit a lognormal distribution to the forward incubation-period distribution because we are relying on existing estimates rather than raw data.
Using this incubation-period distribution, we fit \eref{forwardserial} to the observed serial-interval data by minimizing the negative log-likelihood.
We then calculate the mean forward generation interval using \eref{meangen}.
The 95\% confidence intervals are calculated by taking the estimated variance-covariance matrix for the log-mean and -standard deviation parameters of the log normal distributions and calculating the corresponding variance-covariance for the overall mean using Taylor expansion---this method is also known as the Delta method \citep{oehlert1992note}.
We assume $\rho = 0.75$ throughout based on \cite{sender2021unmitigated}---since we do not have individual-level data on infection and symptom onset times, we expect this parameter to be unidentifiable in practice.
In Supplementary Material, we explore how assumptions about $\rho$ affect inferences of the generation-interval distribution.

@@

\fref{epidemic} summarizes the epidemiological context in the Netherlands during the study period.
The first known Omicron case in the Netherlands was sampled on 19 November 2021 \citep{backer2021omicron}, during a period when COVID-19 incidence was decreasing (\fref{epidemic}A).
As the Omicron variant continued to spread and increase in proportion (\fref{epidemic}B), the number of COVID-19 cases started to increase (\fref{epidemic}A).
Multiplying the proportion of each variant with the number of reported COVID-19 cases further allows us to estimate the epidemiological dynamics of each (\fref{epidemic}C).
The number of COVID-19 cases infected with the Delta variant continued to decrease throughout the study period with time-varying growth rates decreasing from $r \approx -0.01/\mathrm{day}$ to $r \approx -0.09/\mathrm{day}$ by the week of January 16, 2022, and increasing back up to $r \approx -0.04/\mathrm{day}$ by the end of January, 2022 (\fref{epidemic}D).
The number of COVID-19 cases infected with the Omicron variant increased rapidly but decelerated over time with time-varying growth rates decreasing from $r=0.18/\mathrm{day}$ on the week of December 19, 2021, to $r=0.04/\mathrm{day}$ by the end of January, 2022.
These changes in growth rates coincide with the introduction of lockdown on 19 December 2021 \citep{netherlandslockdown} and its relaxation beginning 15 January 2022 \citep{netherlandsrelax,netherlandsopen}.
We note that the growth-rate difference between the Delta and Omicron variants decreased over time.
Hereafter, we use $r=-0.05/\mathrm{day}$ for the Delta variant and $r=0.15/\mathrm{day}$ for the Omicron variant as representative growth rates---these growth rates correspond to the mean growth rates between 1 December 2021 and 2 January 2022, during which the incubation-period data were collected.
We then evaluate the growth-rate effects across $r=-0.1/\mathrm{day}$--$0.0/\mathrm{day}$ for the Delta variant and $r=0.1/\mathrm{day}$--$0.2/\mathrm{day}$ for the Omicron variant as a sensitivity analysis.

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{figure/figure_epidemic.ggp.pdf}
\caption{
\textbf{Epidemic dynamics on the Delta and Omicron variants in the Netherlands between November 2021 and January 2022.}
(A) Daily numbers of reported COVID-19 cases in the Netherlands (points).
The solid line represents the 7-day moving average.
Data are publicly available on \url{https://data.rivm.nl/covid-19/}.
(B) Proportion of SARS-CoV-2 variants detected from the Netherlands. Data are publicly available on \url{https://www.rivm.nl/coronavirus-covid-19/virus/varianten}.
(C) Weekly numbers of COVID-19 cases infected with the Delta (black points) and Omicron (orange triangles) variants are estimated by multiplying the weekly numbers of cases (A) with the proportion of each variant (B).
Solid lines and shaded areas represent fitted lines and corresponding 95\% confidence intervals using generalized additive model.
(D) Estimated growth rates of the Delta (black) and Omicron variants (orange) and their growth-rate differences (purple).
Lines and shaded areas represent medians and corresponding 95\% confidence intervals.
Growth rates are estimated by taking the derivative of the generalized additive model estimates of logged number of cases.
\label{fig:epidemic}
}
\end{figure}

@@

Our corrected estimates of the forward incubation-period distributions further allow us to infer the forward generation-interval distributions.
For illustrative purposes, we first focus on aggregated serial intervals from infectors who developed symptoms during week 50--51 (13--26 December, 2021).
For within-household transmission pairs (\fref{serial}A), the Omicron variant has shorter mean serial interval (3.1 days; 95\% CI: 2.9--3.3 days) than that of the Delta variant (3.7 days; 95\% CI: 3.5--3.8 days).
When we account for growth-rate differences (assuming $r=-0.05/\mathrm{day}$ and $r=0.15/\mathrm{day}$ for the Delta and Omicron variants, respectively), the estimated mean forward generation interval exhibits a slightly larger difference (\fref{serial}B):
3.0 days (95\% CI: 2.7--3.2 days) for the Omicron variant and 3.8 days (95\% CI: 3.7--4.0 days) for the Delta variant.
Across plausible ranges of assumptions about the growth rates of the Delta and Omicron variants, we estimate robust differences in their mean generation intervals (\fref{serial}C--D).
Assuming lower values of the correlation $\rho$ between the incubation period and generation intervals leads to larger differences in the mean generation intervals of the Delta and Omicron variants (Supplementary Figure S1).
In particular, the generation-interval estimates of the Omicron variant are more sensitive to the assumed values of $\rho$ due to faster changes in incidence of infection---for example, changing $\rho$ from 0.85 to 0.5 changes the mean generation-interval estimates for the Omicron variant from 3.1 days (95\% CI: 2.8--3.3 days) to 2.7 days (95\% CI: 2.5--2.9 days).
We cover a wide range of $\rho$ to consider the possibility that our original assumpton ($\rho=0.75$) may under or over-estimate the true $\rho$.
